---
title: "Jug"
author: "Bart Smeets"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

> __Jug is under development and NOT production ready__

## Hello World!

```{r, eval=FALSE}
library(jug)

jug() %>%
  get("/", function(req, res, err){
    "Hello World!"
  }) %>%
  serve_it()
```

```
Serving the jug at http://127.0.0.1:8080
```

## What is Jug?

Jug is a small web development framework for R which relies heavily upon the ```httpuv``` package. It's main focus is to make building APIs for your code as easy as possible.

Jug is not supposed to be either an especially performant nor an uber stable framework. Other tools (and languages) might be more suited for that. Nevertheless, it tries to make available a set of functions which allow you to easily create APIs for your R code. The flexibility of Jug means that, in theory, you could built an extensive web framework with it (but I don't especially recommend it).

## Getting started

For now, Jug is not available on CRAN. To install it, use ```devtools```:

```{r, eval=FALSE}
devtools::install_github("Bart6114/jug")
```

## The Jug instance

Everything starts with a Jug instance. This instance is created by simply calling ```jug()```:

```{r, warning=FALSE, message=FALSE}
library(jug)
jug()
```

Jug is made to work closely with the piping functionality of ```magrittr``` (```%>%```). The rest of the Jug instance is built up by piping the instance through various functions explained below.

## Middleware

In terms of middleware Jug somewhat follows the specification of middleware by ```Express```. In Jug, middleware is a function with access to the __request__ (```req```), __response__ (```res```) and __error__ (```err```) object.

Multiple middlewares can be defined. Order in which the middlewares are added matters. A request will start with being passed through the first middleware added (more specifically the functions specified in it - see next paragraph). It will continue to be passed through the added middlewares until a middleware does not return ```NULL```. Whatever will be passed by that middleware will be set as the response body.

Most middleware will accept a ```func``` or ```...``` argument to which respectively a function or multiple functions can be passed. If multiple functions are passed; passing order will be respected. To each function the ```req```, ```res``` and ```err``` objects will be passed (and thus should accept them). The result of evaluating the function will be used as the response body (if the returned object is not null).

### Method insensitive middleware

The ```use``` function is a method insensitive middleware specifier. While it is method insensitive, it can be bound to a specific path. If the ```path``` argument (accepts a regex string with ```grepl``` setting ```perl=TRUE```) is set to ```NULL``` it also becomes path insensitive and will process _every_ request.

A path insensitive example:

```{r, eval=FALSE}
jug() %>%
  use(path = NULL, function(req, res, err){
    "test 1,2,3!"
    }) %>%
  serve_it()
```

```
$ curl 127.0.0.1:8080/xyz
test 1,2,3!
```
The same example, but path sensitive:

```{r, eval=FALSE}
jug() %>%
  use(path = "/", function(req, res, err){
    "test 1,2,3!"
    }) %>%
  serve_it()
```

```
$ curl 127.0.0.1:8080/xyz
curl: (52) Empty reply from server

$ curl 127.0.0.1:8080
test 1,2,3!
```

Note that in the above error / missing route handling is missing, more on that later.


### Method sensitive middleware

In the same style ass the request method insensitive middleware, there is request method sensitive middleware available. More specifically, you can use the ```get```, ```post```, ```put``` and ```delete``` functions.

This type of middleware is bound to a path using the ```path``` argument. If ```path``` is set to ```NULL``` it will bind to every request to the path, given that it is of the corresponding request type. 

```{r, eval=FALSE}
jug() %>%
  get(path = "/", function(req, res, err){
    "get test 1,2,3!"
    }) %>%
  serve_it()
```

```
$ curl 127.0.0.1:8080
get test 1,2,3!
```

Middlewares are meant to be chained, so to bind different functions to different paths:

```{r, eval=FALSE}
jug() %>%
  get(path = "/", function(req, res, err){
    "get test 1,2,3 on path /"
    }) %>%
  get(path = "/my_path", function(req, res, err){
    "get test 1,2,3 on path /my_path"
    }) %>%
  serve_it()
```

```
$ curl 127.0.0.1:8080
get test 1,2,3 on path /

$ curl 127.0.0.1:8080/my_path
get test 1,2,3 on path /my_path
```

### Including elsewhere defined middleware

You can include middleware that has been defined elsewhere within your current R environment or in a different R file. To do this you can use a ```collector```

Below a ```collector``` is defined locally and ```include```d.

```{r, eval=FALSE}
 collected_mw<-
    collector() %>%
    get("/", function(req,res,err){
      return("test")
    })

  res<-jug() %>%
    include(collected_mw) %>%
    serve_it()
```

However, it is also possible to ```include``` a ```collector``` that is defined in another .R file.

Let's say below is the file ```my_middlewares.R```:

```{r, eval=FALSE}
library(jug)

collected_mw<-
  collector() %>%
  get("/", function(req,res,err){
    return("test2")
  })
```

We can include it as follows:

```{r, eval=FALSE}
res<-jug() %>%
  include(collected_mw, "my_middlewares.R") %>%
  serve_it()
```

## Predefined middleware

### Error handling

A simple error handling middleware (```simple_error_handler```) which catches unbound paths and ```func``` evaluation errors. If no custom error handler is implemented I suggest you add this to your Jug instance.

```{r, eval=FALSE}
jug() %>%
  simple_error_handler() %>%
  serve_it()
```

```
$ curl 127.0.0.1:8080
<!DOCTYPE html>
<html lang="en">
  <head>
    <meta charset="utf-8">
    <title>Not found</title>
  </head>
  <body>
    <p>No handler bound to path</p>
  </body>
</html>
```
If you want to implement your own custom error handling just have a look at the ```jug::simple_error_handler()``` function.

### Easily using your own functions

The main reason Jug was created is to easily allow access to your own custom R functions. The convenience function ```decorate``` is built with this in mind.

If you ```decorate``` your own function it will translate all arguments passed in the query string of the request as arguments to your function. It will also pass all headers to the function as arguments. Note that for the passed headers they will be capitalized and prefixed by ```HTTP_``` as per ```httpuv```'s internals. 

If your function does not accept a ```...``` argument, all query/header parameters that are nor explicitly requested by your function are dropped. If your function requests a ```req```, ```res``` or ```err``` argument (or ```...```) the corresponding objects will be passed. 

```{r, eval=FALSE}
say_hello<-function(name){paste("hello",name,"!")}

jug() %>%
  get("/", decorate(say_hello)) %>%
  serve_it()
```

```
$ curl 127.0.0.1:8080/?name=Bart
hello Bart !
```

### Static file server

The ```serve_static_file``` middleware allows for serving static files.

```{r, eval=FALSE}
jug() %>%
  serve_static_files() %>%
  serve_it()
```

The default root directory is the one returned by ```getwd()``` but can be specified by providing a ```root_path``` argument to the ```serve_static_files``` middleware. It transforms a bare ```/``` path to ```index.html```.

## The request, response and error objects

### Request (```req```) object

The ```req``` object contains the request specifications. It has different attributes:

- ```req$params``` a list of the query parameters contained in the URL
- ```req$path``` the requested path
- ```req$method``` the request method
- ```req$raw``` the raw request object as passsed by ```httpuv```
- ```req$post_data``` the post data (has some limitations for now...)

It has the following functions attached to it:

- ```req$get_header(key)``` returns the value associated to the specified key in the request (no need to worry about the ```HTTP_``` prefix)
- ```req$attach(key, value)``` attach a variable to ```req$params```

### Response (```res```) object

The ```res``` object contains the response specifications. It has different attributes:

- ```res$headers``` a named list of the set headers
- ```res$status``` the status of the response (defaults to 200)
- ```res$body``` the body of the response (is automatically set by the content of the not ```NULL``` returning middleware)

It also has a set of functions:

- ```res$set_header(key, value)``` set a custom header
- ```res$content_type(type)``` set your own content type (MIME)
- ```res$set_status(status)``` set the status of the response
- ```res$text(body)``` to explicitely set the body of the response
- ```res$json(obj)``` converts an object to JSON, sets it as the body and set the correct content type

### Error (```err```) object

The ```err``` object contains a list of errors, accessible through ```err$errrors```. You can add an error to this list by calling ```err$set(error)```. The error will be converted to a character.

Have a look at the ```simple_error_handler``` middleware function see how you can implement a custom error handler.

## URL dispatching 

The path parameter in the ```get```, ```post```, ... functions is intepreted as a regex pattern.

If there are named capture groups in the path definition, they will be attached to the ```req$params``` object. For example the pattern ```/test/(?<id>.*)/(?<id2>.*)``` will result in the variables ```id``` and ```id2``` (with their respective values) being bound to the ```req$params``` object.

If a path pattern is not started with a start of string ```^``` regex token __or__ ended with an end of string token ```$```, these will be explicitely inserted at respectively the beginning and end of the path pattern specification. For example the path pattern ```/``` will be converted to ```^/$```.

## Serving the Jug

Simple call ```serve_it()``` at the end of your piping chain (see Hello World! example).

## Further examples

[Minimal CRUD TODO app](https://github.com/Bart6114/jug-crud-example)
