---
title: "Jug"
author: "Bart Smeets"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Hello World!

```{r, eval=FALSE}
library(jug)

jug() %>%
  get("/", function(req, res, err){
    "Hello World!"
  }) %>%
  serve_it()
```
```
Serving the jug at http://127.0.0.1:8080
```

## What is Jug?

Jug is a webdevelopment framework for R which relies heavily upon the ```httpuv``` package.

Jug is not supposed to be either an especially performant nor an uber stable framework. Other tools (and languages) might be more suited for that. Nevertheless, it tries to make available a set of functions which allow you to easily create APIs for your R code. The flexibility of Jug means that, in theory, you could built an extensive web framework with it (but I don't especially recommend it).

## The basics

### The Jug instance

Everything starts with a Jug instance. This instance is created by simply calling ```jug()```:

```{r, warning=FALSE, message=FALSE}
library(jug)
jug()
```

Jug is made to work closely with the piping functionality of ```magrittr``` (```%>%```). The rest of the Jug instance is built up by piping the instance through various functions explained below.

### Middleware

In terms of middleware Jug somewhat follows the specification of middleware by ```Express```. In Jug, middleware is a function with access to the __request__ (```req```), __response__ (```res```) and __error__ (```err```) object.

Multiple middlewares can be defined. Order in which the middlewares are added matters. A request will start with being passed through the first middleware added. It will continue to be passed through the added middlewares until a middleware does not return ```NULL```. Whatever will be passed by the middleware will be set as the response body.

#### Method insensitive middleware

The ```use``` function is a method insensitive middleware specifier. While it is method insensitive, it can be bound to a specific path. If the ```path``` argument is set to ```NULL``` it also becomes path insensitive and will process _every_ request.

A path insensitive example:

```{r, eval=FALSE}
jug() %>%
  use(path = NULL, function(req, res, err){
    "test 1,2,3!"
    }) %>%
  serve_it()
```

```
$ curl 127.0.0.1:8080/xyz
test 1,2,3!
```
The same example, but path sensitive:

```{r, eval=FALSE}
jug() %>%
  use(path = "/", function(req, res, err){
    "test 1,2,3!"
    }) %>%
  serve_it()
```

```
$ curl 127.0.0.1:8080/xyz
curl: (52) Empty reply from server

$ curl 127.0.0.1:8080
test 1,2,3!
```

Note that in the above error / missing route handling is missing, more on that later.


### Predefined middleware

### Easily using your own functions

### Serving


